---
title: "BioOutputs"
output:
  md_document:
    md_extensions: -markdown_in_html_blocks
    toc: yes
---

-----

# BioOutputs

This package contains common R scripts I use in my day to day data analysis of biological data. The scripts are primarily for plotting and visualisation, with some data organisation thrown in as well. 

-----

## Gallery


```{r, results='asis', echo=FALSE}
plots <- list.files("./figs", pattern="*.png")
plots <- paste0("./figs", '/', plots)
 

cat('<table border="1">
  <tr>
    <td align="center" style="vertical-align:top" height="200"><a href="#bires">Bi-Results</a><img src=',plots[1],' height="150" width="330"/></td>
    <td align="center" style="vertical-align:top" height="200"><a href="#corrp">Correlation Plot</a><img src=',plots[2],' height="150" width="330"/></td>
    <td style="vertical-align:top" align="center" height="200"><a href="#freq">Frequency Table</a><img src=',plots[3],' height="50" width="330"/></td>
  </tr>
<tr>
    <td align="center" style="vertical-align:top" height="200"><a href="#mod">Module Plot</a><img src=',plots[4],' height="150" width="330"/></td>
    <td align="center" style="vertical-align:top" height="200"><a href="#volc">Volcano Plot</a><img src=',plots[5],' height="150" width="330"/></td>
  </tr>
</table>')


```

-----

## Required packages 
  [`dplyr`](https://dplyr.tidyverse.org)
  [`ggplot2`](https://ggplot2.tidyverse.org)
  [`ggrepel`](https://cran.r-project.org/web/packages/ggrepel/vignettes/ggrepel.html)

-----

## Install Package

First install devtools to allow installation from gitub and any other required packages. 
```{r, eval=FALSE}
install.packages("devtools")
library("devtools")
```

```{r}
library(devtools)
library(knitr)
```

Now install the BioOutputs package.
```{r, eval=FALSE}
install_github("KatrionaGoldmann/BioOutputs")
library("BioOutputs")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library("devtools")
library(roxygen2)

#Initialize the package
setwd("~/Documents/Git/")

# Load/update the package
package.name="BioOutputs"
setwd(paste("~/Documents/Git/", package.name, sep=""))
document()
```

-----

<a id="corrp"></a>

## bio_corr 

Create a correlation plot. Taken from kassambara/ggpubr just changed the default arguments. 

So we can use the classic example with the *mtcars* data frames:

```{r}
kable(head(mtcars))
```

```{r bio_corr}
bio_corr(mtcars, "qsec", "wt")
```

-----

<a id="freq"></a>

## bio_frequency

The *bio_frequency()* function generates a frequency table from factor or character vector columns in a data frame. This has the following arguments:

Argument           | 
-------------------|----------------------------------------------------
data         | A data frame containing columns to be counted  
columns           | Column names or indices to be counted in data  
freq.percent     | Whether the table should include frequency counts, percentages or both (options = c("freq", "percent", "both")). Default="both"  
include.na         | Include NA values (options are TRUE/FALSE, default=TRUE)  
remove.vars         | Character vector of variables not to be included in the counts (e.g. remove.vars = c("") remove blanks from the count)



Then if we want to see the breakdown of, say, the gear column in mtcars we can apply:

```{r}
kable(bio_frequency(mtcars, "gear"))
```

And if wanted we can remove one variable from the table. This is useful if we have unknowns or the likes.

```{r}
kable(bio_frequency(mtcars, "gear", remove.vars=c("5")))
```

-----

<a id="volc"></a>

## bio_volcano

This function generates a volcano plot from a top table using ggplot. The function contains many parameters, use `?bio_volcano` to interogate. 

Lets look at the leukemia data set

```{r, message=FALSE, warning=FALSE, eval=FALSE}
BiocManager::install("leukemiasEset", version = "3.8")
```

```{r, message=FALSE, warning=FALSE}
library(leukemiasEset)
library(limma)
```

```{r, message=FALSE, warning=FALSE}
data(leukemiasEset)
ourData <- leukemiasEset[, leukemiasEset$LeukemiaType %in% c("ALL", "NoL")]
ourData$LeukemiaType <- factor(ourData$LeukemiaType)

design <- model.matrix(~ ourData$LeukemiaType)
fit <- lmFit(ourData, design)
fit <- eBayes(fit)
toptable <- topTable(fit)
toptable$pvalue = toptable$P.Value
```

```{r bio_volcano}
bio_volcano(toptable, fc.col="logFC", label.row.indices=1:10, main="leukemia", add.lines=FALSE)
```


-----

<a id="bires"></a>

## bio_bires 

This function colours data according to whether it is below or above a defined plane. The plane is plotted as a line and data can be output as either a line or markers/points. 

```{r bires}
data(beavers)
df.plane = beaver1
df.data = beaver2
df.plane$temp <- df.plane$temp + 0.5
bio_bires(x="time", y="temp", df.data, x.plane="time", y.plane="temp", df.plane)
```

## bio_mods

This function splits expression data into customisable modules and averages over catagories in a given variable. In this example we will look at two Li modules and a custom one I made up. 

```{r}
load("/Users/phd/Downloads/Complete_web_data_rld.rdata")
cos.mods <- readRDS("/Users/PhD/Downloads/Module_list_KG.rds")
```

```{r bires}
mod_list <- all.mod.list[c("M3", "M47.0", "M156.0")]
mod_list[["custom MS4A1"]] <- cos.mods$gene[cos.mods$group == "MS4A1"]

exp = rld.syn
meta <- rld.metadata.syn
mean.var = "Pathotype"
cols = NULL
cluster.rows=FALSE
keep(mod_list, exp, meta, cluster.rows, cols, mean.var, sure=T)
bio_mods(exp=exp, mod.list=mod_list, meta=meta, mean.var = "Pathotype", show.names=FALSE, cluster.rows=FALSE)

```

