---
title: "BioOutputs"
output:
  md_document:
    md_extensions: -markdown_in_html_blocks
---

-----


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, eval=TRUE)
```

# BioOutputs

This package contains common R scripts I use in my day to day data analysis of biological data. The scripts are primarily for plotting and visualisation, with some data organisation thrown in as well. 

-----

## Gallery


```{r, results='asis', echo=FALSE}
plots <- list.files("./figs", pattern="*.png")
plots <- paste0("./figs", '/', plots)
 

cat('<table border="1">
  <tr>
    <td align="center" style="vertical-align:top" height="200"><a href="#bires">Bi-Results</a><img src=',plots[1],' height="150" width="330"/></td>
    <td align="center" style="vertical-align:top" height="200"><a href="#corrp">Correlation Plot</a><img src=',plots[2],' height="150" width="330"/></td>
    <td style="vertical-align:top" align="center" height="200"><a href="#freq">Frequency Table</a><img src=',plots[4],' height="50" width="330"/></td>
  </tr>
	<tr>
    <td align="center" style="vertical-align:top" height="200"><a href="#mods">Module Plot</a><img src=',plots[5],' height="150" width="330"/></td>
    <td align="center" style="vertical-align:top" height="200"><a href="#tgcomp">Treatment Group Comparisons</a><img src=',plots[7],' height="150" width="330"/></td>
    <td align="center" style="vertical-align:top" height="200"><a href="#volc">Volcano Plot</a><img src=',plots[8],' height="150" width="330"/></td>
  </tr>
 	<tr>
		<td align="center" style="vertical-align:top" height="200"><a href="#geneids">Switching Gene Ids</a></td>
		<td align="center" style="vertical-align:top" height="200"><a href="#fcp">Fold Change Heatmap</a><img src=',plots[3],' height="150" width="330"/></td>
    <td align="center" style="vertical-align:top" height="200"><a href="#boxp">Significance Boxplots</a><img src=',plots[6],' height="150" width="330"/></td>
  </tr>
</table>')


```

-----

## Required packages 
  [`dplyr`](https://dplyr.tidyverse.org)
  [`ggplot2`](https://ggplot2.tidyverse.org)
  [`ggrepel`](https://cran.r-project.org/web/packages/ggrepel/vignettes/ggrepel.html)
  [`ComplexHeatmap`](http://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html)
  [`RColorBrewer`](https://www.rdocumentation.org/packages/RColorBrewer/versions/1.1-2/topics/RColorBrewer)
  [`Rmisc`](https://www.rdocumentation.org/packages/Rmisc/versions/1.5)
  [`ggpubr`](https://cran.r-project.org/web/packages/ggpubr/index.html)
  [`gtools`](https://cran.r-project.org/web/packages/gtools/index.html)
  [`grid`](https://www.rdocumentation.org/packages/grid/versions/3.5.2)
  [`pBrackets`](https://cran.r-project.org/web/packages/pBrackets/index.html)
  [`biomaRt`](https://bioconductor.org/packages/release/bioc/html/biomaRt.html)


-----

## Install Package

First install devtools to allow installation from gitub and any other required packages. 
```{r, eval=FALSE}
install.packages("devtools")
library("devtools")
```

```{r, echo=FALSE}
library(devtools)
library(knitr)
library(cowplot)
```

```{r}
library(devtools)
library(knitr)
```

Now install the BioOutputs package.
```{r, eval=FALSE}
install_github("KatrionaGoldmann/BioOutputs")
library("BioOutputs")
```

```{r, echo=FALSE, eval=FALSE}
library(BioOutputs)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide', eval=TRUE}
library("devtools")
library(roxygen2)

package.name="BioOutputs"
# Load/update the package (desktop)
setwd(paste("~/Documents/Scripts/Git/", package.name, sep=""))
#Initialize the package (laptop)
#setwd(paste("~/Documents/Git/", package.name, sep=""))

document()
```

-----

<a id="corrp"></a>

## bio_corr 


Create a correlation plot. Taken from kassambara/ggpubr just changed the default arguments. 

So we can use the classic example with the *mtcars* data frames:

```{r}
kable(head(mtcars))
```

```{r bio_corr}
bio_corr(mtcars, "qsec", "wt")
```

-----

<a id="freq"></a>


## bio_frequency

The *bio_frequency()* function generates a frequency table from factor or character vector columns in a data frame. This has the following arguments:

Argument           | 
-------------------|----------------------------------------------------
data         | A data frame containing columns to be counted  
columns           | Column names or indices to be counted in data  
freq.percent     | Whether the table should include frequency counts, percentages or both (options = c("freq", "percent", "both")). Default="both"  
include.na         | Include NA values (options are TRUE/FALSE, default=TRUE)  
remove.vars         | Character vector of variables not to be included in the counts (e.g. remove.vars = c("") remove blanks from the count)



Then if we want to see the breakdown of, say, the gear column in mtcars we can apply:

```{r}
kable(bio_frequency(mtcars, "gear"))
```

And if wanted we can remove one variable from the table. This is useful if we have unknowns or the likes.

```{r}
kable(bio_frequency(mtcars, "gear", remove.vars=c("5")))
```

-----


<a id="volc"></a>


## bio_volcano

This function generates a volcano plot from a top table using ggplot. The function contains many parameters, use `?bio_volcano` to interogate. 

Argument           | 
-------------------|----------------------------------------------------
toptable        | A data frame containing p value and fold change columns for parameters compared across multiple groups. The p value column should be named "pvalue".  
fc.col           | The column name which stores the fold change. Should be in the log2 format (default="log2FC")
padj.col     | The column which contains adjusted p-values. If NULL adjusted pvalues will be calculated 
padj.method	         | correction method. Options include: c("holm", "hochberg", "hommel", "bonferroni", "BH", "BY", "fdr", "none"). Default="fdr
padj.cutoff         | The cutoff for adjusted pvalues. This adds a horizontal line of significance (default=NULL)
fc.cutoff         | The log2(fold change) significance cut-off (default=1)
marker.colour         | Character vector of four colours to map to the volcano plot. In the order non-significanct, fold-change significant, pvalue significant, significant in fold-change and pvalues (default=c("grey60", "olivedrab", "salmon", "darkturquoise"))
label.p.cutoff         | The cutoff for adjusted pvalues for labelling (default=NULL). Not recommended if many significant rows.
label.row.indices         | Indices of rows to be labelled (default=NULL)
label.colour         | Colour of labels (default="black")
legend.labs         | A character vector for theThe legend label names (default=c("Not Significant", "FC>fc.cutoff", "Padj<padj.cutoff", "FC>fc.cutoff& Padj<padj.cutoff"))
add.lines         | Whether to add dashed lines at fc.cutoff and padj.cutoff (default=TRUE)
line.colour         | The color of dashed significance lines (default="grey14")
main         | Plot title
xlims, ylims         | The plot limits



Lets look at the ALS patients carrying the C9ORF72 data set

<!-- ```{r, message=FALSE, warning=FALSE, eval=FALSE} -->
<!-- BiocManager::install("leukemiasEset", version = "3.8") -->
<!-- ``` -->

<!-- ```{r, message=FALSE, warning=FALSE} -->
<!-- library(leukemiasEset) -->
<!-- library(limma) -->
<!-- ``` -->

<!-- ```{r, message=FALSE, warning=FALSE} -->
<!-- data(leukemiasEset) -->
<!-- ourData <- leukemiasEset[, leukemiasEset$LeukemiaType %in% c("ALL", "NoL")] -->
<!-- ourData$LeukemiaType <- factor(ourData$LeukemiaType) -->

<!-- design <- model.matrix(~ ourData$LeukemiaType) -->
<!-- fit <- lmFit(ourData, design) -->
<!-- fit <- eBayes(fit) -->
<!-- toptable <- topTable(fit) -->
<!-- toptable$pvalue = toptable$P.Value -->
<!-- ``` -->

```{r}
toptable <- read.table("https://gist.githubusercontent.com/stephenturner/806e31fce55a8b7175af/raw/1a507c4c3f9f1baaa3a69187223ff3d3050628d4/results.txt", sep="", header=T)
rownames(toptable) = toptable$Gene
```


```{r bio_volcano}
bio_volcano(toptable, fc.col="log2FoldChange", label.row.indices=1:10, main="ALS patients carrying the C9ORF72", add.lines=TRUE)
```

Lets look at the leukemia data set

```{r, message=FALSE, warning=FALSE, eval=FALSE}
BiocManager::install("leukemiasEset", version = "3.8")
```

```{r, message=FALSE, warning=FALSE}
library(leukemiasEset)
library(limma)
```

```{r, message=FALSE, warning=FALSE}
data(leukemiasEset)
ourData <- leukemiasEset[, leukemiasEset$LeukemiaType %in% c("ALL", "NoL")]
ourData$LeukemiaType <- factor(ourData$LeukemiaType)

design <- model.matrix(~ ourData$LeukemiaType)
fit <- lmFit(ourData, design)
fit <- eBayes(fit)
toptable <- topTable(fit)
toptable$pvalue = toptable$P.Value
```

```{r}
bio_volcano(toptable, fc.col="logFC", label.row.indices=1:10, main="leukemia", add.lines=FALSE)
```


-----

<a id="bires"></a>

## bio_bires 

This function colours data according to whether it is below or above a defined plane. The plane is plotted as a line and data can be output as either a line or markers/points. 

Argument           | 
-------------------|----------------------------------------------------
x         | x column name in data
y         | y column name in data
df.data         | Data frame containing x and y columns
x.plane         | column name for the x axis in df.plane
y.plane         | column name for the y axis in df.plane
df.plane         | Date frame modelling the plane
stepwise         | logical whether to plot the cutoff plane as stepwise or smoothed
colours         | colour vector for higher, lower and plane values (default=c("green", "red", "grey) respectively)
inc.equal         | logical whethere points on the line should be counted as above (dafault=TRUE)
labels         | label for the markers (default=c("above", "below"))
type         | type of plot for data (options include point (dafault), line, stepwise)


Lets look at some beaver temperature data. 


```{r bires}
data(beavers)
df.plane = beaver1
df.data = beaver2
df.plane$temp <- df.plane$temp + 0.5
bio_bires(x="time", y="temp", df.data, x.plane="time", y.plane="temp", df.plane)
```

<a id="mods"></a>

## bio_mods

This function splits expression data into customisable modules and averages over catagories in a given variable. 

Argument           | 
-------------------|----------------------------------------------------
exp         | data frame containing the expression data
mod.list         | A list of modules. Each element contains the list of genes for a modules. The gene names must match the rownames in the exp dataframe.
meta         | Dataframe where each column contains an annotation/tract for samples in the heatmap. The order of samples in meta must match that of exp.
cluster.rows         | The method to use for clustering of rows
cols         | Chacter vector, or named vector to fix the order, defining the colours of each mean.var group.
main         | Title of heatmap
show.names         | Show row names. Logical.
mean.subjects         | Logical to determine whether to add a row for the mean value for all subjects in a group
split.var         | Character defining the meta column to average (mean) over.


In this example we will look at two Li modules and a custom one I made up. 

```{r, echo=FALSE, eval=TRUE}
#load("/Users/phd/Downloads/Complete_web_data_rld.rdata")
load("~/V:/Katriona/ShinyDocker/CurrentWebsite/DIY/Complete_web_data_rld.rdata")
#cos.mods <- readRDS("/Users/PhD/Downloads/Module_list_KG.rds")
cos.mods <- readRDS("~/V:/Katriona/PEAC_analysis/GeneSets/Module_list_KG.rds")

mod_list <- all.mod.list[c("M3", "M47.0", "M156.0")]
mod_list[["custom MS4A1"]] <- as.character(cos.mods$gene[cos.mods$group == "MS4A1"])

names(mod_list) = c(paste("mod", 1:4))
```

```{r mods, eval=TRUE}
exp = rld.syn
meta <- rld.metadata.syn

bio_mods(exp=exp, mod.list=mod_list, meta=meta, mean.var = "Pathotype", cluster.rows = FALSE)
```

<a id="tgcomp"></a>

## bio_treatmentGroups

This function plots a line graph comparing two treatment groups over time. 

Argument           | 
-------------------|----------------------------------------------------
df         | Data frame containing the data for both groups
group.col         | Column name which corresponding to the group of values in df
y.col         | Column name corresponding to the y-axis values in df
x.col         | Column name corresponding to the x-axis values in df
id.col         | Column name which corresponds to subject ID
main         | Title of pot
cols         | Character vector for colours of lines
p.col         | Colour of p-value text

Lets look at kidney disease survival in this example from the survival package. 

```{r, warning=FALSE}
library(survival)
data(kidney)

df <- kidney
df <- df[df$time < 150, ]

bio_treatmentGroups(df, group.col="status", y.col="frail", x.col="time")
```

<a id="geneids"></a>

## bio_geneid

This function switches gene (or snp) ids using biomart

Argument           | 
-------------------|----------------------------------------------------
IDs         | list of the Ids you want to convert
IDFrom         | What format these IDs are in (default Ensembl)
IDTo         | What format you want the IDs converted to (default gene names)
mart         | The biomart to use. Typically, for humans you will want ensembl (default). Alternatives can be found at listEnsembl()
dataset         | you want to use. To see the different datasets available within a biomaRt you can e.g. do: mart = useEnsembl('ENSEMBL_MART_ENSEMBL'), followed by listDatasets(mart).
attributes         | list of variables you want output

For example genes TNF and A1BG:

```{r}
IDs = c("TNF", "A1BG")
kable(bio_geneid(IDs, IDFrom='hgnc_symbol', IDTo = 'ensembl_transcript_id'))
```


<a id="fcp"></a>

## bio_fc_heatmap

This function exports a complex heatmap looking at the expression of different modules which can be customised. 


Argument           | 
-------------------|----------------------------------------------------
exp         | Expression Data
var         | Vector classing samples by variables
prefix         | Prefix to Heatmap titles
stars         | whether pvales should be written as numeric or start (default=FALSE)
overlay         | pvalues on fold change Heatmap or besidde (default = TRUE)
logp         | Whether or not to log the pvalues
...         | Other parameters to pass to Complex Heatmap

```{r, eval=FALSE}
exp = rld.syn
meta <- rld.metadata.syn

bio_fc_heatmap(exp=exp, var=meta$Pathotype)
```


<a id="boxp"></a>

## bio_boxplots

Creates boxplots showing the significance between groups.

Argument           | 
-------------------|----------------------------------------------------
data         | Data frame containing columns x and y
x, y         | x and y variable names for drawing.
p.cutoff         | plot p-value if above p.cutoff threshold. To include all comparisons set as NULL.
stars         | Logical. Whether significance shown as numeric or stars
method         | a character string indicating which method to be used for comparing means.c("t.test", "wilcox.test")
star.vals         | a list of arguments to pass to the function symnum for symbolic number coding of p-values. For example, the dafault is symnum.args <- list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1), symbols = c('****', '***', '**', '*', 'ns')). In other words, we use the following convention for symbols indicating statistical significance: ns: p > 0.05; *: p <= 0.05; **: p <= 0.01; ***: p <= 0.001; ****: p <= 0.0001

Lets look at the iris data:

```{r}
bio_boxplots(iris, x="Species", y= "Sepal.Width", p.cutoff = 0.0001)
bio_boxplots(iris, x="Species", y= "Sepal.Width", NULL, stars=TRUE)
```
