---
title: "BioOutputs"
output:
  md_document:
    md_extensions: -markdown_in_html_blocks
    toc: yes
---

-----


# BioOutputs

This package contains common R scripts I use in my day to day data analysis of biological data. The scripts are primarily for plotting and visualisation, with some data organisation thrown in as well. 

-----

## Gallery


```{r, results='asis', echo=FALSE}
plots <- list.files("./figs", pattern="*.png")
plots <- paste0("./figs", '/', plots)
 

cat('<table border="1">
  <tr>
    <td align="center" style="vertical-align:top" height="200"><a href="#bires">Bi-Results</a><img src=',plots[1],' height="150" width="330"/></td>
    <td align="center" style="vertical-align:top" height="200"><a href="#corrp">Correlation Plot</a><img src=',plots[2],' height="150" width="330"/></td>
    <td style="vertical-align:top" align="center" height="200"><a href="#freq">Frequency Table</a><img src=',plots[4],' height="50" width="330"/></td>
  </tr>
	<tr>
    <td align="center" style="vertical-align:top" height="200"><a href="#mods">Module Plot</a><img src=',plots[5],' height="150" width="330"/></td>
    <td align="center" style="vertical-align:top" height="200"><a href="#tgcomp">Treatment Group Comparisons</a><img src=',plots[6],' height="150" width="330"/></td>
    <td align="center" style="vertical-align:top" height="200"><a href="#volc">Volcano Plot</a><img src=',plots[7],' height="150" width="330"/></td>
  </tr>
 	<tr>
		<td align="center" style="vertical-align:top" height="200"><a href="#geneids">Switching Gene Ids</a></td>
		<td align="center" style="vertical-align:top" height="200"><a href="#fcp">Fold Change Heatmap</a><img src=',plots[3],' height="150" width="330"/></td>
    <td align="center" style="vertical-align:top" height="200"></td>
  </tr>
</table>')


```

-----

## Required packages 
  [`dplyr`](https://dplyr.tidyverse.org)
  [`ggplot2`](https://ggplot2.tidyverse.org)
  [`ggrepel`](https://cran.r-project.org/web/packages/ggrepel/vignettes/ggrepel.html)
  [`ComplexHeatmap`](http://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html)
  [`RColorBrewer`](https://www.rdocumentation.org/packages/RColorBrewer/versions/1.1-2/topics/RColorBrewer)
  [`Rmisc`](https://www.rdocumentation.org/packages/Rmisc/versions/1.5)
  [`ggpubr`](https://cran.r-project.org/web/packages/ggpubr/index.html)
  [`grid`](https://www.rdocumentation.org/packages/grid/versions/3.5.2)
  [`pBrackets`](https://cran.r-project.org/web/packages/pBrackets/index.html)
  [`biomaRt`](https://bioconductor.org/packages/release/bioc/html/biomaRt.html)


-----

## Install Package

First install devtools to allow installation from gitub and any other required packages. 
```{r, eval=FALSE}
install.packages("devtools")
library("devtools")
```

```{r, echo=FALSE}
library(devtools)
library(knitr)
library(cowplot)
```

```{r}
library(devtools)
library(knitr)
```

Now install the BioOutputs package.
```{r, eval=FALSE}
install_github("KatrionaGoldmann/BioOutputs")
library("BioOutputs")
```

```{r, echo=FALSE}
library(BioOutputs)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide', eval=FALSE}
library("devtools")
library(roxygen2)

#Initialize the package (laptop)
# package.name="BioOutputs"
# setwd(paste("~/Documents/Git/", package.name, sep=""))


# Load/update the package (desktop)
package.name="BioOutputs"
setwd(paste("~/Documents/Scripts/Git/", package.name, sep=""))
document()
```

-----

<a id="corrp"></a>

## bio_corr 


Create a correlation plot. Taken from kassambara/ggpubr just changed the default arguments. 

So we can use the classic example with the *mtcars* data frames:

```{r}
kable(head(mtcars))
```

```{r bio_corr}
bio_corr(mtcars, "qsec", "wt")
```

-----

<a id="freq"></a>


## bio_frequency

The *bio_frequency()* function generates a frequency table from factor or character vector columns in a data frame. This has the following arguments:

Argument           | 
-------------------|----------------------------------------------------
data         | A data frame containing columns to be counted  
columns           | Column names or indices to be counted in data  
freq.percent     | Whether the table should include frequency counts, percentages or both (options = c("freq", "percent", "both")). Default="both"  
include.na         | Include NA values (options are TRUE/FALSE, default=TRUE)  
remove.vars         | Character vector of variables not to be included in the counts (e.g. remove.vars = c("") remove blanks from the count)



Then if we want to see the breakdown of, say, the gear column in mtcars we can apply:

```{r}
kable(bio_frequency(mtcars, "gear"))
```

And if wanted we can remove one variable from the table. This is useful if we have unknowns or the likes.

```{r}
kable(bio_frequency(mtcars, "gear", remove.vars=c("5")))
```

-----


<a id="volc"></a>


## bio_volcano

This function generates a volcano plot from a top table using ggplot. The function contains many parameters, use `?bio_volcano` to interogate. 

Lets look at the ALS patients carrying the C9ORF72 data set

<!-- ```{r, message=FALSE, warning=FALSE, eval=FALSE} -->
<!-- BiocManager::install("leukemiasEset", version = "3.8") -->
<!-- ``` -->

<!-- ```{r, message=FALSE, warning=FALSE} -->
<!-- library(leukemiasEset) -->
<!-- library(limma) -->
<!-- ``` -->

<!-- ```{r, message=FALSE, warning=FALSE} -->
<!-- data(leukemiasEset) -->
<!-- ourData <- leukemiasEset[, leukemiasEset$LeukemiaType %in% c("ALL", "NoL")] -->
<!-- ourData$LeukemiaType <- factor(ourData$LeukemiaType) -->

<!-- design <- model.matrix(~ ourData$LeukemiaType) -->
<!-- fit <- lmFit(ourData, design) -->
<!-- fit <- eBayes(fit) -->
<!-- toptable <- topTable(fit) -->
<!-- toptable$pvalue = toptable$P.Value -->
<!-- ``` -->

```{r}
toptable <- read.table("https://gist.githubusercontent.com/stephenturner/806e31fce55a8b7175af/raw/1a507c4c3f9f1baaa3a69187223ff3d3050628d4/results.txt", sep="", header=T)
rownames(toptable) = toptable$Gene
```


```{r bio_volcano}
bio_volcano(toptable, fc.col="log2FoldChange", label.row.indices=1:10, main="ALS patients carrying the C9ORF72", add.lines=TRUE)
```

Lets look at the leukemia data set

```{r, message=FALSE, warning=FALSE, eval=FALSE}
BiocManager::install("leukemiasEset", version = "3.8")
```

```{r, message=FALSE, warning=FALSE}
library(leukemiasEset)
library(limma)
```

```{r, message=FALSE, warning=FALSE}
data(leukemiasEset)
ourData <- leukemiasEset[, leukemiasEset$LeukemiaType %in% c("ALL", "NoL")]
ourData$LeukemiaType <- factor(ourData$LeukemiaType)

design <- model.matrix(~ ourData$LeukemiaType)
fit <- lmFit(ourData, design)
fit <- eBayes(fit)
toptable <- topTable(fit)
toptable$pvalue = toptable$P.Value
```

```{r}
bio_volcano(toptable, fc.col="logFC", label.row.indices=1:10, main="leukemia", add.lines=FALSE)
```


-----

<a id="bires"></a>

## bio_bires 

This function colours data according to whether it is below or above a defined plane. The plane is plotted as a line and data can be output as either a line or markers/points. 

```{r bires}
data(beavers)
df.plane = beaver1
df.data = beaver2
df.plane$temp <- df.plane$temp + 0.5
bio_bires(x="time", y="temp", df.data, x.plane="time", y.plane="temp", df.plane)
```

<a id="mods"></a>

## bio_mods

This function splits expression data into customisable modules and averages over catagories in a given variable. In this example we will look at two Li modules and a custom one I made up. 

```{r, echo=FALSE}
#load("/Users/phd/Downloads/Complete_web_data_rld.rdata")
load("~/V:/Katriona/ShinyDocker/CurrentWebsite/DIY/Complete_web_data_rld.rdata")
#cos.mods <- readRDS("/Users/PhD/Downloads/Module_list_KG.rds")
cos.mods <- readRDS("~/V:/Katriona/PEAC_analysis/GeneSets/Module_list_KG.rds")

mod_list <- all.mod.list[c("M3", "M47.0", "M156.0")]
mod_list[["custom MS4A1"]] <- as.character(cos.mods$gene[cos.mods$group == "MS4A1"])

names(mod_list) = c(paste("mod", 1:4))
```

```{r mods}
exp = rld.syn
meta <- rld.metadata.syn

bio_mods(exp=exp, mod.list=mod_list, meta=meta, mean.var = "Pathotype", cluster.rows = FALSE)
```

<a id="tgcomp"></a>

## bio_treatmentGroups

```{r, warning=FALSE}
library(survival)
data(kidney)

df <- kidney
df <- df[df$time < 150, ]

bio_treatmentGroups(df, group.col="status", y.col="frail", x.col="time")
```

<a id="geneids"></a>

## bio_geneid

```{r}
IDs = c("TNF", "A1BG")
kable(bio_geneid(IDs, IDFrom='hgnc_symbol', IDTo = 'ensembl_transcript_id'))
```


<a id="fcp"></a>

## bio_fc_heatmap

```{r}
exp = rld.syn
meta <- rld.metadata.syn

bio_fc_heatmap(exp=exp, var=meta$Pathtye)
```
